---
title: "eDNA Data Analysis"
documentclass: article
classoption: a4paper
output: word_document
fig_height: 50%
fig_width: 50%
geometry: margin=0.6in
---

# eDNA Data Analysis Workshop 2022

## Learning Outcomes:

1. Students are able to appropriately analyse data from an eDNA field sampling campaign;
2. Students are able to interpret and understand statistical outputs from the R package 'vegan'.

```{r set up, results='hide'}
library(vegan)
library(flextable)
eDNA2022 <- read.csv(file = "eDNA2022.csv")
eDNA2022$Site <- as.factor(eDNA2022$Site)
```

## List of all the species and their common names

```{r}
altnames <- 
  c("MosqFish", "Goby", "Mullet", "Bream", "BnjoFrog", "GToadlet", "MlrdDuck",
    "SpotDove", "SwampHen", "Crake", "WFHeron", "Lorikeet", "Cattle", "Sheep",
    "Mouse", "Rat", "HydroidB", "HydroidO", "BwnJelly", "MoonJely", "Dog")
names_table <- 
  data.frame(Taxonomic = gsub("."," ",colnames(eDNA2022[,4:24]), fixed = T),
             Common=c("Mosquito Fish","Bridled Goby","Gray Mullet","Black Bream",
                      "Western Banjo Frog", "Gunther's Toadlet", "Mallard Duck",
                      "Spotted Turtle Dove", "Purple Swamphen", "Spotted Crake",
                      "White Faced Heron", "Rainbow Lorikeet", "Cattle", "Sheep",
                      "Mouse", "Rat", "HydroidB", "HydroidO", "Brown Jellyfish",
                      "Moon Jellyfish", "Domestic Dog"),
             Abbrviated=altnames)
flextable(names_table,cwidth = c(3,2,2))
```

```{r generate all species data, results='hide'}
allSpecies <- eDNA2022[,4:24]
colnames(allSpecies) <- altnames
```

## Detrended correspondence analysis - all species

```{r}
AF_dca_all <- decorana(allSpecies)
AF_dca_all
```
```{r decorana plot, fig.height=5, fig.width=5}
plot(AF_dca_all)
```

# Ordinations

Next we will construct our ordination – a mechanism to visualise the similarity
matrix. There are two different types of ordination that we will use –
**Multi-Dimensional-Scaling** (nmMDS) and **Principal Coordinates Analysis**
(PCO). If you have previously used PCA – Principal Component Analysis - you may
be familiar with ordination methods. PCA is useful for environmental datasets –
these are represented by a similarity matrix derived via use of Euclidean
Distance as a measure of similarity. PCA is more suited to environmental
variables because of the type of data – there are no large blocks of zeros
(usually) and it is no longer necessary to select a similarity measure that
ignores joint absences. When looking at ordination plots in general you should
note that sample points that are spatially located together share common
characteristics and sample points that are distantly located from one another
share less characteristics. This holds whether we are working with environmental
variables (pH, EC, moisture content, heavy metals *etc*.) or species abundances.

The main difference between MDS and PCO relates to the underlying use of the
resemblance matrix.

**MDS** takes the similarity measures calculated by Jaccard or Bray Curtis and
ranks them in order. Thus pairs are samples are ranked according to highest to
lowest similarity and are thus plotted in multidimensional space.

**PCO** takes the value of underlying measure of similarity and plots pairs of
samples in multi-dimensional space starting with the pairs of samples with
highest level of similarity and working through the pairs of samples.

For both types of ordinations the software attempts to preserve the similarity
of each pair of samples –- while this is feasible in multidimensional space, in
order for us to view the ordination we see this in 2 or 3 dimensions -– thus the
software attempts to preserve as much as is feasible of the sample pair
similarity as calculated –- but in reality we need to consider how much the 2-
dimensional view is a true representation of the data set. In nmMDS we use the
*stress* value to evaluate this, in PCO we can plot multiple axes, *i.e*. axis 1
*vs*. axis 2, axis 2 *vs*. axis 3 and so on. In reality much of the variability
is often explained in the first two axes (*i.e*. 2 dimensions) and we will limit
our analysis to these axes for this dataset.

## Non-metric multidimensional scaling (nmMDS)

```{r run nmds all jaccard, results='hide'}
AF_nmds_all <- metaMDS(allSpecies, distance = "jaccard")
```

```{r show mds all jaccard}
AF_nmds_all
```

## nmds again with Bray
```{r run nmds Bray allspecies, results='hide'}
AF_nmds_all_B <- metaMDS(allSpecies) # uses Bray by default
```

```{r show nmds Bray allspecies, results='hold'}
AF_nmds_all_B
```

This plot mimics the nmmds plot in Deirdre's handout (note that the sign of dimension MDS2 is reversed)

```{r plot nmds all Bray, fig.height=6, fig.width=6}
plot(AF_nmds_all$points, pch=c(rep(21:25, 2),21)[eDNA2022$Site], 
     xlim = c(-1.6,2.2), ylim = c(1.2,-1.2), 
     bg=seq(1,11)[eDNA2022$Site], 
     cex = 1.4, main = "All species", col.main = "steelblue")
text(jitter(AF_nmds_all$points[,1], factor = 500),
     jitter(AF_nmds_all$points[,2], factor=50), labels=eDNA2022$SiteID, cex=1, 
     col=c("black","grey40","grey70",
           rainbow(4,v=0.75,end=0.25),
           rainbow(4,v=0.75,start=0.45,end=0.8))[eDNA2022$Site])
text(AF_nmds_all, display = "species", col = "#80008080", font=3, cex = 1.1)
```

## Analysis with just the aquatic species

These are: *Gambusia holbrooki, Arenigobius bifrenatus, Mugil cephalus, Acanthopagrus butcheri, Limnodynastes dorsalis, Pseudophryne guentheri, Blackfordia polytentaculata, Obelia bidentata, Phyllorhiza punctata, Aurelia sp.*

```{r subset to just aquatic, paged.print=FALSE, results='hold'}
fish <- allSpecies[,c(1:6,17:20)]
head(fish)
```

### for interest look at the dissimilarity matrix

```{r dissim matrix aquat bray}
AF_diss_aquat <- vegdist(fish, 
                           distance = "bray")
options(max.print=2000)
print(AF_diss_aquat, digits = 2)
```

```{r run nmds aquat Bray, results='hide'}
AF_nmds_aquat <- metaMDS(fish, trymax = 500, 
                           distance = "bray")
```
```{r show nmds aquat Bray}
AF_nmds_aquat
```
```{r plot nmds aquat Bray, fig.height=5, fig.width=5}
plot(AF_nmds_aquat, type="p",display = "sites", cex = 1.2)
text(AF_nmds_aquat, display = "species", col = "dodgerblue", cex = 0.8)
```

### alternative to vegan plots
```{r alt1 to vegan nmds plots, fig.height=5, fig.width=5}
palette(c("black","grey80","grey92",
          rainbow(4,s=0.5,end=0.25),rainbow(4,s=0.5,start=0.45,end=0.8)))
plot(AF_nmds_aquat$points, pch=c(rep(21:25, 2),21)[eDNA2022$Site], 
     xlim = c(-1.2,1.8), 
     bg = seq(1,11)[eDNA2022$Site], 
     cex = 1.4, col.main = "steelblue", main = "Just aquatic species")
text(jitter(AF_nmds_aquat$points[,1], factor = 500),
     jitter(AF_nmds_aquat$points[,2], factor=50), 
     labels=eDNA2022$SiteID, cex=1, 
     col=c("black","grey40","grey70",
           rainbow(4,v=0.75,end=0.25),
           rainbow(4,v=0.75,start=0.45,end=0.8))[eDNA2022$Site])
text(AF_nmds_aquat, display = "species", col = "grey33", font=3, cex = 0.9)
```

## Principal Coordinates Analysis

In the 'vegan' package we use Weighted classical multidimensional scaling, also
known as *Weighted Principal Coordinates Analysis*.

```{r}
AF_pco_all <- wcmdscale(vegdist(allSpecies), eig=TRUE)
```
```{r basic pco plot, fig.height=5, fig.width=5, fig.cap="Basic PCO plot based on all species."}
plot(AF_pco_all)
```

```{r better pco plot using base R, fig.width=6, fig.height=6}
palette(c("black","grey60","grey42",
          rainbow(4,v=0.75,end=0.25),rainbow(4,v=0.75,start=0.45,end=0.8)))
plot(AF_pco_all$points[,1:2], xlim = c(0.4,-0.6), ylim = c(0.4,-0.6), 
     pch=3, cex = 0.5, col = "grey")
text(AF_pco_all$points[,1:2],labels = eDNA2022$SiteID,
     col=seq(1,11)[eDNA2022$Site])
```

```{r calculate pco scores but not sure if we need this}
AF_pco_scores <- scores(AF_pco_all)
```

In the 'vegan' package, the adonis2() function implements Permutational Multivariate Analysis of Variance Using Distance Matrices

```{r}
adonis2(eDNA2022[,4:24] ~ eDNA2022$Site, data = eDNA2022)
```

adonis2(Site ~ , data = eDNA2022)
